---
title: "Duwaertis and Mattis RNA-seq human iPSC lines"
author: "Tara Friedrich"
date: "2/19/2019"
output: html_document
---

```{r setup, include=FALSE}
library(pheatmap)
library(edgeR)
library(DESeq2)
library(biomaRt)
library(ggplot2)
library(topGO)
require(plyr)
library(GOplot)
library("org.Hs.eg.db")
sessionInfo()
```


Read in file with read counts.

```{r, include=FALSE}
filename1 = "~/Downloads/readcounts1.txt"
filename2 = "~/Downloads/readcounts2.txt"
filename3 = "~/Downloads/readcounts3.txt"

counts1 = read.table(filename1, header = T, stringsAsFactors = F) 
counts2 = read.table(filename2, header = T, stringsAsFactors = F) 
counts3 = read.table(filename3, header = T, stringsAsFactors = F) 

counts = cbind(counts1, counts2, counts3)
dim(counts)
rownames(counts) = counts1$Geneid

```



Remove metadata (Gene, coordinates, etc)
```{r}
counts_matrix = data.matrix(counts[,grep ("bam", colnames(counts))])
#colnames(counts_matrix) = sub("_1_algn_Aligned", "", matrix(unlist(strsplit(colnames(counts_matrix), "\\.")), byrow =T, ncol=6)[,3])
colnames(counts_matrix) = matrix(unlist(strsplit(colnames(counts_matrix), "algn.")), byrow =T, ncol=3)[,2]
#rename to reflect useful labels 
key1 = read.table("~/Downloads/key.txt", header = T, stringsAsFactors = F, fill=T, sep="\t")
head(key1)
key2 = read.table("~/Downloads/Mattis_Duwaerts_metadata.txt", header = T, stringsAsFactors = F, fill=T, sep="\t")
head(key2)

renamesample=list()
for (x in 1:nrow(key1  )){
  oldname = paste0(sub(".fastq", "", key1$FILE.NAME[x], "_"), "_")
  #print (key1$CELL.TYPE[x])  #check to make sure every sample is endoderm 
  newname = paste(key1$FULL.SAMPLE.ID[x], key1$CELL.TYPE[x], key1$DISEASE.STATUS[x], "CD", sep="_")
  renamesample[oldname] = newname
}
for (x in 1:nrow(key2)){
  oldname = paste0(sub("-", ".", sub(".fq.gz", "", matrix(unlist(strsplit( key2$Location[x], "/")), byrow=T, ncol=7)[,7])), "_")
  #print (key2$cell.type[x])  #check to make sure every sample is endoderm 
  newname = paste(key2$Name[x], key2$cell.of.origin[x], key2$Genotype[x], "AM", sep="_")
  renamesample[oldname] = newname
}

colnames(counts_matrix) = unlist(lapply (colnames(counts_matrix), function(x) renamesample[[x]]))
colnames(counts_matrix) = sub ("NASH pt", "NASH", colnames(counts_matrix))
colnames(counts_matrix) = sub ("wild-type", "HEALTHY", colnames(counts_matrix))

if (colnames(counts_matrix)[21] == colnames(counts_matrix)[22]) {
  colnames(counts_matrix)[21] = "CW10027_iPSC-PBMC_HEALTHY_CDa"
  colnames(counts_matrix)[22] = "CW10027_iPSC-PBMC_HEALTHY_CDb"
}
colnames(counts_matrix)[grep ("WBC", colnames(counts_matrix))] = c("CW10054_iPSC-PBMC_HEALTHY_AM", "CW10007_iPSC-PBMC_NASH_AM" )

```

Filter out genes that are too highly expressed or lowly expressed. 
```{r}
keep = rowMeans(counts_matrix) > 5 & rowMeans(counts_matrix) < 5000 
counts_matrix_filtered = counts_matrix[keep,]
```
Take the RPKM.
```{r}
y <- DGEList(counts=counts_matrix_filtered)
fpkm_matrix = rpkm (y, gene.length=counts1[keep,"Length"], log = FALSE)
```

Dendrogram showing relationships using fpkm. Mattis' PBMCs do not cluster with Caroline's even though two of them come from the same patient.  CW10027 samples do not cluster together.   CW10020, CW10022 and CW10054 were listed as outliers prior to analysis but they don't seem to be behaving oddly.  
```{r}
#par(cex=1,font=3)
hc2 <- hclust(stats::dist(t(fpkm_matrix), method="minkowski"), "ward.D2")
#par(mar=c(10, 4, 4, 10), pty='s')
plot(hc2, lwd=4, lty=1, col="black", col.lab="red", xlab="Samples", main="Samples clustered by FPKM using Ward's clustering & Euclidean distance")

```

label each sample by three conditions
1) cell origin 2) disease status 3) patient source
```{r}
Condition1 = factor(matrix (unlist(strsplit(colnames(counts_matrix_filtered), "_")), byrow=T, ncol=4)[,2], levels = c("iPSC-PBMC", "fibroblast"))
Condition2 = factor(matrix (unlist(strsplit(colnames(counts_matrix_filtered), "_")), byrow=T, ncol=4)[,3], levels = c("HEALTHY","NASH", "NAFL"))
Condition3 = factor(matrix (unlist(strsplit(colnames(counts_matrix_filtered), "_")), byrow=T, ncol=4)[,1])
```

PCA log transformation
```{r}
cold <- data.frame("Condition1"=Condition1) 
rownames(cold) = colnames(counts_matrix_filtered)
cold

ddsMat <- DESeq2::DESeqDataSetFromMatrix(countData=counts_matrix_filtered, colData=cold, design=~Condition1 );
colnames(ddsMat) <- colnames(counts_matrix_filtered)
rld <- DESeq2::rlog(ddsMat) # LOG TRANSFORMED.

```

plot PCA
```{r}
par(mfrow=c(2,1))
pca <- prcomp(t(assay(rld)))

d <- data.frame(pca$x, Condition1, Condition2,  name=colnames(rld)) # returns PC1, PC2, ... PC8... etc
ppp <- ggplot(d, aes_string(x="PC1", y="PC2",shape="Condition1", color="Condition2"))
ppp <- ppp + geom_point(size=6)
plot(ppp)

d <- data.frame(pca$x, group=Condition1, Condition3,  name=colnames(rld)) # returns PC1, PC2, ... PC8... etc
ppp <- ggplot(d, aes_string(x="PC1", y="PC2",shape="Condition1", color="Condition3"))
ppp <- ppp + geom_point(size=6)
plot(ppp)
```

Some of the related samples are messing up clustering. 

Heatmap to see how cells cluster by cell type or experimental condition. Does it agree with dendrogram which uses different clustering methods. 
```{r}
HEATMAP_HEIGHT <- 45 # inchescol
HEATMAP_WIDTH  <- 15
annotation <- data.frame(
  Condition1= Condition1, Condition2=Condition2)
rownames(annotation) <- colnames(fpkm_matrix)
pheatmap(fpkm_matrix,
         , width=HEATMAP_WIDTH, height=HEATMAP_HEIGHT
         , scale="row"  # actually scale each row
         , clustering_distance_rows="correlation" # or: euclidean
         , clustering_distance_cols="correlation"
         , clustering_method="complete"
         , display_numbers=FALSE
         , border_color="#00000000" # note: the last two digits here are the alpha channel/ transparency
         , show_rownames=FALSE, show_colnames = TRUE
         , annotation = annotation
         , fontsize_row=3.5)
```


Fit the model and estimate dispersion value. Baseline is PBMC.  
```{r}
design_matrix <- model.matrix( ~ Condition1  )
colnames(design_matrix)
design_matrix
y_for_DE <- DGEList(counts=counts_matrix_filtered)
y_for_DE <- calcNormFactors(y_for_DE)
y_for_DE = estimateDisp(y_for_DE, design=design_matrix) 
plotBCV(y_for_DE)
fit <- glmFit(y_for_DE, design=design_matrix, dispersion=y_for_DE$trended.dispersion)
```
Pretty high estimated dispersion values. 


Fibroblast vs PBMC 
```{r}
lrt <- glmLRT(fit, coef = "Condition1fibroblast")
topTags(lrt)
smoothScatter(x=lrt$table$logFC, y=-log10(lrt$table$PValue))
```
More genes appear to be down-regulated. 



```{r}
lrt_table = data.frame(lrt$table, p.adjust(lrt$table$PValue, method = "BH"))
colnames(lrt_table) = c(colnames(lrt$table), "adjPVal_BH")
hist(lrt_table$PValue)
```

Format output into excel sheet. 
```{r}
genes = rownames(counts_matrix_filtered)
ensembl = useEnsembl(biomart="ensembl", dataset="hsapiens_gene_ensembl")
head(listAttributes(ensembl), 30)
hgnc_swissprot <- getBM(attributes=c('ensembl_gene_id','hgnc_symbol','description'),filters = 'ensembl_gene_id',  values = genes , mart = ensembl)

index = match (genes, hgnc_swissprot$ensembl_gene_id)
results = data.frame(hgnc_swissprot[index,], counts_matrix_filtered, fpkm_matrix, lrt_table)
colnames(results) = c(
                    colnames(hgnc_swissprot[index,])
                  , unlist(lapply (colnames(counts_matrix_filtered), function(x) paste(x, "counts", sep="_")))
                  , unlist(lapply (colnames(fpkm_matrix), function(x) paste(x, "fpkm", sep="_")))
                  , unlist(lapply (colnames(lrt_table), function(x) paste(x, "_fibroblast_vs_pbmc_baseline", sep="_")))
                  )
curr_date = format(Sys.time(), "%a_%b_%d_%H_hrs_%M_min_%S_sec_%Y")
write.table(results, paste0("~/Box Sync/laptop_folders/duwaerts_mattis_de_", curr_date, ".xls"), row.names=rownames(counts_matrix_filtered), col.names=NA, sep="\t")

```


Heatmap of DE genes
```{r}
mean_values = 2^((fit$coefficients) %*% t(design_matrix) )
colnames(mean_values) = colnames (counts_matrix_filtered)

de_mean_values = mean_values[rownames(mean_values) %in%  results[which(results[,"adjPVal_BH__fibroblast_vs_pbmc_baseline"] <=0.05),"ensembl_gene_id"], ]
pheatmap(de_mean_values,
         , width=HEATMAP_WIDTH, height=HEATMAP_HEIGHT
         , scale="row"  # actually scale each row
         #, clustering_distance_rows="correlation" # or: euclidean
         , clustering_distance_cols="euclidean"
         , clustering_method="complete"
         , display_numbers=FALSE
        , border_color="#00000000" # note: the last two digits here are the alpha channel/ transparency
         , show_rownames=FALSE, show_colnames = TRUE
         #, annotation = annotation
         , fontsize_row=3.5)

de_fpkm_values = results[which(results[,"adjPVal_BH__fibroblast_vs_pbmc_baseline"] <=0.05), grep ("_fpkm", colnames(results))]
pheatmap(de_fpkm_values,
         , width=HEATMAP_WIDTH, height=HEATMAP_HEIGHT
         , scale="row"  # actually scale each row
         #, clustering_distance_rows="correlation" # or: euclidean
         , clustering_distance_cols="euclidean"
         , clustering_method="complete"
         , display_numbers=FALSE
        , border_color="#00000000" # note: the last two digits here are the alpha channel/ transparency
         , show_rownames=FALSE, show_colnames = TRUE
         #, annotation = annotation
         , fontsize_row=3.5)
```

```{r}
#setup gene DE for GOplot package later
GOplot_genes = results [grep ("__fibroblast_vs_pbmc_baseline",colnames(results))]
colnames(GOplot_genes) = c("logFC", "logCPM", "LR", "Pvalue", "BH")
GOplot_genes["ID"] = results[,"ensembl_gene_id"]
colnames(GOplot_genes)
length(unique(GOplot_genes$ID))
GOplot_genes = GOplot_genes[, c("ID", "logFC")]

geneIDs = results[,"ensembl_gene_id"]
inUniverse = geneIDs %in%  results[,"ensembl_gene_id"]
inSelection = geneIDs %in% results[which(results[,"adjPVal_BH__fibroblast_vs_pbmc_baseline"] <=0.05),"ensembl_gene_id"]
alg <- factor( as.integer( inSelection[inUniverse] ) )
names(alg) <- geneIDs[inUniverse]
head(alg)

onts = c( "MF", "BP", "CC" )
tab = as.list(onts)
names(tab) = onts
GOplot_terms=list()
  for(i in 1:3) {
    ## prepare data
    tgd <- new( "topGOdata", ontology=onts[i], allGenes = alg, nodeSize=5,
                 annot=annFUN.org, mapping="org.Hs.eg.db", ID = "ensembl" )

    ## run tests
    resultTopGO.elim <- runTest(tgd, algorithm = "elim", statistic = "Fisher" )
    resultTopGO.classic <- runTest(tgd, algorithm = "classic", statistic = "Fisher" )

    ## look at results
    tab[[i]] <- GenTable( tgd, Fisher.elim = resultTopGO.elim, 
        Fisher.classic = resultTopGO.classic,
        orderBy = "Fisher.classic" , topNodes = 200)

    #convert list to dataframe  
    allGO = genesInTerm(tgd, tab[[i]][,"GO.ID"])
    #allGO_genes = ldply(allGO,data.frame)[,2]
    #length(unique(allGO_genes))
    #GOplot_genes = GOplot_genes[GOplot_genes$ID %in% allGO_genes,]
    test = ldply (lapply(allGO, function(x) paste(x, collapse=",")), data.frame)
    #match(test$.id, tab[[i]][, "GO.ID"]) #check to make sure they rows match 
    GOplot_terms[[i]] = data.frame(onts[i], tab[[i]], test, stringsAsFactors = FALSE)
    GOplot_terms[[i]][, "Fisher.elim"] = as.numeric(as.character(GOplot_terms[[i]][, "Fisher.elim"]))
    GOplot_terms[[i]][, "Fisher.elim.BH"] = p.adjust(GOplot_terms[[i]][, "Fisher.elim"], method="BH")
    GOplot_terms[[i]] = GOplot_terms[[i]][, c("onts.i.", "GO.ID", "Term", "Fisher.elim.BH", "X..i..")]
    colnames(GOplot_terms[[i]]) = c("Category", "ID", "Term",  "adj_pval", "Genes")
  }

topGOResults <- rbind.fill(tab)
#write.csv(topGOResults, file = "~/Box Sync/laptop_folders//topGOResults.csv")
topGOResults2 <- rbind.fill(GOplot_terms)
#write.csv(topGOResults2, file = "~/Box Sync/laptop_folders//topGOResults2.csv")
                  
curr_date = format(Sys.time(), "%a_%b_%d_%H_hrs_%M_min_%S_sec_%Y")
write.table(results, paste0("~/Box Sync/laptop_folders/duwaerts_mattis_goresults_", curr_date, ".xls"),  col.names=NA, sep="\t")

circ <- circle_dat(topGOResults2, data.frame(GOplot_genes))
GOBar(circ)
GOBar(circ, display='multiple')
GOBubble(circ, labels = 1.5)
GOCircle(circ)

```