---
Title: "Willenbring RNA-seq to identify TFs"
author: "Tara Friedrich"
date: "5/7/2019"
output: html_document
---

```{r }
library(pheatmap)
library(edgeR)
library(DESeq2)
library(biomaRt)
library(ggplot2)
library(topGO)
require(plyr)
library(GOplot)
library("org.Hs.eg.db")
sessionInfo()
```

Read in files with read counts.  
```{r}
filename1 = "~/Downloads/readcounts_GSE108315.txt"
filename2 = "~/Downloads/readcounts_GSE55552.txt"
counts1 = read.table(filename1, header = T, stringsAsFactors = F) 
counts2 = read.table(filename2, header = T, stringsAsFactors = F)
```

Remove metadata (Gene, coordinates, etc)
```{r}
counts_matrix = data.frame(counts1[,grep ("bam", colnames(counts1))], counts2[,grep ("bam", colnames(counts2))])
colnames(counts_matrix) = sub("GSE55552..algn.GSE55552_sra_", "", sub(".sra_algn_Aligned.sortedByCoord.out.bam", "", sub("GSE108315..algn.GSE108315_sra_", "", colnames(counts_matrix))))
rownames(counts_matrix) = counts1$Geneid

#rename columns/sample to reflect useful labels 
key = read.table("~/Downloads/key_SRR_to_name.txt", header = T, stringsAsFactors = F, fill=T, sep="\t")
head(key)

renamesample=list()
studyOrder=list()
groupOrder=list()
for (x in 1:nrow(key)){
  oldname = key$Accession_Code[x]
  newname = key$Name[x]
  renamesample[oldname] = newname
  studyOrder[newname] = key$Study[x]
  groupOrder[newname] = key$Group[x]
}

colnames(counts_matrix) = unlist(lapply (colnames(counts_matrix), function(x) renamesample[[x]]))
StudyStatus = unlist(lapply (colnames(counts_matrix), function(x) studyOrder[[x]]))
GroupStatus = unlist(lapply (colnames(counts_matrix), function(x) groupOrder[[x]]))
#shorten column names
colnames(counts_matrix) = sub("Donor-derived ducts ", "", sub("Host-derived ducts ", "", sub("Hepatocyte ", "", colnames(counts_matrix))))
SampleID = colnames(counts_matrix)
```
Filter out genes that are too highly expressed or lowly expressed. Highly expressed genes are usually mitochondrial. 
```{r}
length(rowMeans(counts_matrix) < 5000 )
dim(counts_matrix)
length(which(rowMeans(counts_matrix) > 5000 ))
keep = rowMeans(counts_matrix) > 5 
counts_matrix_filtered = counts_matrix[keep,]
```
Take the RPKM.
```{r}
y <- DGEList(counts=counts_matrix_filtered)
fpkm_matrix = rpkm (y, gene.length=counts1[keep,"Length"], log = FALSE)
```
Dendrogram showing relationships using fpkm. 
```{r}
par(cex=.5,font=2)
hc2 <- hclust(stats::dist(t(fpkm_matrix), method="minkowski"), "ward.D2")
par(mar=c(.2,.2,.2,.2), pty='s')
plot(hc2, lwd=4, lty=1, col="black", col.lab="red", xlab="Samples", main="Samples clustered by FPKM using Ward's clustering & Euclidean distance")
```


log transformation for PCA plot 
```{r}
cold <- data.frame("GroupStatus"=GroupStatus, "StudyStatus"=StudyStatus) 
rownames(cold) = colnames(counts_matrix_filtered)
cold

ddsMat <- DESeq2::DESeqDataSetFromMatrix(countData=counts_matrix_filtered, colData=cold, design=~GroupStatus );
colnames(ddsMat) <- colnames(counts_matrix_filtered)
rld <- DESeq2::rlog(ddsMat) # LOG TRANSFORMED.
```

run PCA to explore structure in the data in lower dimensions.  
```{r}
par(mfrow=c(2,1))
pca <- prcomp(t(assay(rld)))
```


Plot PC1 against PC2 using variance stabilizing transformation. 

```{r}
d <- data.frame(pca$x, group=GroupStatus, SampleID, StudyStatus,  name=colnames(rld)) # returns PC1, PC2, ... PC8... etc
ppp <- ggplot(d, aes_string(x="PC1", y="PC2",shape="GroupStatus", color="StudyStatus"))
ppp <- ppp + geom_point(size=3) + geom_text(aes(label=SampleID), nudge_y=6)
plot(ppp)
```  

```{r}
HEATMAP_HEIGHT <- 45 # inchescol
HEATMAP_WIDTH  <- 15
annotation <- data.frame(
  StudyStatus= StudyStatus, GroupStatus=GroupStatus)
rownames(annotation) <- colnames(fpkm_matrix)
pheatmap(fpkm_matrix,
         , width=HEATMAP_WIDTH, height=HEATMAP_HEIGHT
         , scale="row"  # actually scale each row
         , clustering_distance_rows="correlation" # or: euclidean
         , clustering_distance_cols="correlation"
         , clustering_method="complete"
         , display_numbers=FALSE
         , border_color="#00000000" # note: the last two digits here are the alpha channel/ transparency
         , show_rownames=FALSE, show_colnames = TRUE
         , annotation = annotation
         , fontsize_row=3.5)
```

Fit the model and estimate dispersion value. Baseline is PBMC.  
```{r}
design_matrix <- model.matrix( ~ GroupStatus + StudyStatus  )
colnames(design_matrix)
design_matrix
y_for_DE <- DGEList(counts=counts_matrix_filtered)
y_for_DE <- calcNormFactors(y_for_DE)
y_for_DE = estimateDisp(y_for_DE, design=design_matrix) 
plotBCV(y_for_DE)
fit <- glmFit(y_for_DE, design=design_matrix, dispersion=y_for_DE$trended.dispersion)
```


```{r}
GroupStatus1 = relevel (factor(GroupStatus[which(StudyStatus==0)]), ref="C")
design_matrix1 <- model.matrix( ~GroupStatus1 )
colnames(design_matrix1)
design_matrix1
counts_matrix_filtered1 = counts_matrix_filtered[,colnames(counts_matrix_filtered)[which(StudyStatus==0)]]


y_for_DE1 <- DGEList(counts=counts_matrix_filtered1)
y_for_DE1 <- calcNormFactors(y_for_DE1)
y_for_DE1 = estimateDisp(y_for_DE1, design=design_matrix1) 
plotBCV(y_for_DE1)
fit1 <- glmFit(y_for_DE1, design=design_matrix1, dispersion=y_for_DE1$trended.dispersion)
```

```{r}
lrt1 <- glmLRT(fit1, coef = "GroupStatus1A")
topTags(lrt1)
smoothScatter(x=lrt1$table$logFC, y=-log10(lrt1$table$PValue))
lrt2 <- glmLRT(fit1, coef = "GroupStatus1B")
topTags(lrt2)
smoothScatter(x=lrt2$table$logFC, y=-log10(lrt2$table$PValue))
```

Histogram should skew left.  
```{r}
lrt_table1 = data.frame(lrt1$table, p.adjust(lrt1$table$PValue, method = "BH"))
colnames(lrt_table1) = c(colnames(lrt1$table), "adjPVal_BH")
hist(lrt_table1$PValue, main="distribution of p-values") 
lrt_table2 = data.frame(lrt2$table, p.adjust(lrt2$table$PValue, method = "BH"))
colnames(lrt_table2) = c(colnames(lrt2$table), "adjPVal_BH")
hist(lrt_table2$PValue, main="distribution of p-values") 
```

Format output into excel sheet. 
```{r}
fpkm_matrix1 = fpkm_matrix[,colnames(counts_matrix_filtered)[which(StudyStatus==0)]]

genes = rownames(counts_matrix_filtered1)
ensembl = useEnsembl(biomart="ensembl", dataset="mmusculus_gene_ensembl",host="www.ensembl.org")
#head(listAttributes(ensembl), 30)
hgnc_swissprot <- getBM(attributes=c('ensembl_gene_id','external_gene_name','description'),filters = 'ensembl_gene_id',  values = genes , mart = ensembl)

index = match (genes, hgnc_swissprot$ensembl_gene_id)
results = data.frame(hgnc_swissprot[index,], counts_matrix_filtered1, fpkm_matrix1, lrt_table1,lrt_table2)
colnames(results) = c(
                    colnames(hgnc_swissprot[index,])
                  , unlist(lapply (colnames(counts_matrix_filtered1), function(x) paste(x, "counts", sep="_")))
                  , unlist(lapply (colnames(fpkm_matrix1), function(x) paste(x, "fpkm", sep="_")))
                  , unlist(lapply (colnames(lrt_table1), function(x) paste(x, "_A_vs_C_baseline", sep="_")))
                  , unlist(lapply (colnames(lrt_table2), function(x) paste(x, "_B_vs_C_baseline", sep="_")))
                  )
curr_date = format(Sys.time(), "%a_%b_%d_%H_hrs_%M_min_%S_sec_%Y")
write.table(results, paste0("~/Downloads/willenbring_", "study1", "_de_", curr_date, ".xls"), row.names=rownames(counts_matrix_filtered), col.names=NA, sep="\t")

```


