---
title: "2018_09_12_Chang_Weng"
output: html_document
---
```{r}
library(pheatmap)
library(edgeR)
library(DESeq2)
library(biomaRt)
library(ggplot2)
sessionInfo()
```

Friday, September 7, 2018 
WT73M_readcount               Bred WT
WT58M_readcount               Bred WT
WT973M_readcount             Bred WT
KO87M_readcount                Bred KO
KO56M_readcount                Bred KO
KO972M_readcount              Bred KO
WT112M_readcount             Virus WT
WT113M_readcount             Virus WT
WT114M_readcount             Virus WT
KO116M_readcount              Virus KO
KO117M_readcount              Virus KO
KO119M_readcount              Virus KO



```{r }
#library(readxl)
setwd("~/Documents/chang/")
#originaldata = read_excel("gene_list/gene.description.xls", sheet="gene.description") #disappointed this didn't work
                         #need to download the developer version to see if that works
#devtools::install_version("readxl", version = "1.0.0")
#https://github.com/tidyverse/readxl/issues/466

```
Read in read counts for each sample taken from gene.description.xls
```{r}
filename = "readcounts.txt"
counts1 = read.table(filename, header = T, stringsAsFactors = F) 
rownames(counts1) = counts1$geneID

```
Rename samples according to Yun's instructions
```{r}
counts2 = counts1[,grep("readcount", colnames(counts1))]
sample_names = list(
  "WT73M_readcount" = "Bred_WT_rep1"
, "WT58M_readcount" = "Bred_WT_rep2"
, "WT973M_readcount" = "Bred_WT_rep3"
, "KO87M_readcount" = "Bred_KO_rep1"
, "KO56M_readcount" = "Bred_KO_rep2"
, "KO972M_readcount" = "Bred_KO_rep3"
, "WT112M_readcount" = "Virus_WT_rep1"
, "WT113M_readcount" = "Virus_WT_rep2"
, "WT114M_readcount" = "Virus_WT_rep3"
, "KO116M_readcount" = "Virus_KO_rep1"
, "KO117M_readcount" = "Virus_KO_rep2"
, "KO119M_readcount" = "Virus_KO_rep3"
)
colnames(counts2) = unlist(lapply (colnames(counts2), function(x) sample_names[[x]]))
colnames(counts2)
```
Normally ilter out genes with low and high counts ( high counts might be mitochondrial genes)
```{r}
#data in table has probably already been filtered so keep all rows in this case
#keep = rowMeans(counts2) > 5 & rowMeans(counts2) < 5000 
#keep = seq(1, nrow(counts2))
#if adjust p-values are NA then they probably removed that gene when testing for DE
keep = !(is.na(counts1$padj.KO_bredvsWT_bred.) | is.na(counts1$padj.KO_virusvsWT_virus.))
counts3 = counts2[keep,]
dim(counts2)
dim(counts3)
```

Take the log (RPKM)
```{r}
y <- DGEList(counts=counts3)
#fpkm_log_matrix = rpkm (y, gene.length=counts1[keep,"GeneLength"], log = TRUE)
fpkm_log_matrix = rpkm (y, gene.length=counts1[keep,"GeneLength"], log = FALSE)

```
Dendrogram 
```{r}
# 
#par(cex=1,font=3)
hc2 <- hclust(stats::dist(t(fpkm_log_matrix), method="minkowski"), "ward.D2")
#par(mar=c(10, 4, 4, 10), pty='s')
plot(hc2, lwd=4, lty=1, col="black", col.lab="red", xlab="Samples", main="Samples clustered by FPKM using Ward's clustering & Euclidean distance")
```
```{r}
Condition1 = factor(matrix (unlist(strsplit(colnames(counts3), "_")), byrow=T, ncol=3)[,1], levels = c("Bred", "Virus"))
Condition2 = factor(matrix (unlist(strsplit(colnames(counts3), "_")), byrow=T, ncol=3)[,2], levels = c("KO", "WT"))
Condition2 = relevel(Condition2, ref="WT")
cold <- data.frame("Condition1"=Condition1, "Condition2"=Condition2) 
rownames(cold) = colnames(counts3)
cold

ddsMat <- DESeq2::DESeqDataSetFromMatrix(countData=counts3, colData=cold, design=~Condition1 + Condition2);
colnames(ddsMat) <- colnames(counts3)
rld <- DESeq2::rlog(ddsMat) # LOG TRANSFORMED.
#rld <- DESeq2::rlog(ddsMat, blind=blind_status) 
```

```{r}
pca              <- prcomp(t(assay(rld)))
#barplot(allPer.vec, main="PCA: % variance explained by each component\n" , ylim=c(0,100), names.arg=allPerLabels.vec)

d <- data.frame(pca$x, group=Condition1, Condition2, name=colnames(rld)) # returns PC1, PC2, ... PC8... etc
ppp <- ggplot(d, aes_string(x="PC1", y="PC2", shape="Condition2", color="group"))
ppp <- ppp + geom_point(size=6)
plot(ppp)
```
Heatmap to see how cells cluster by cell type or experimental condition. 
```{r}
HEATMAP_HEIGHT <- 45 # inchescol
HEATMAP_WIDTH  <- 15
annotation <- data.frame(
  Condition1= Condition1, Condition2=Condition2)
rownames(annotation) <- colnames(fpkm_log_matrix)
pheatmap(fpkm_log_matrix,
         , width=HEATMAP_WIDTH, height=HEATMAP_HEIGHT
         , scale="row"  # actually scale each row
         , clustering_distance_rows="correlation" # or: euclidean
         , clustering_distance_cols="correlation"
         , clustering_method="complete"
         , display_numbers=FALSE
         , border_color="#00000000" # note: the last two digits here are the alpha channel/ transparency
         , show_rownames=FALSE, show_colnames = TRUE
         , annotation = annotation
         , fontsize_row=3.5)
```

Virus: Fit the model and estimate dispersion value.
```{r}
Virus_cols = grep ("Virus", colnames(counts3))
design_matrix_virus <- model.matrix( ~ Condition2[Virus_cols])
colnames(design_matrix_virus)
design_matrix_virus
y_for_DE_virus <- DGEList(counts=counts3[, Virus_cols])
y_for_DE_virus <- calcNormFactors(y_for_DE_virus)
y_for_DE_virus = estimateDisp(y_for_DE_virus, design=design_matrix_virus) 
plotBCV(y_for_DE_virus)
fit_virus <- glmFit(y_for_DE_virus, design=design_matrix_virus, dispersion=y_for_DE_virus$trended.dispersion)
```

Bred: Fit the model and estimate dispersion value.
```{r}
Bred_cols = grep ("Bred", colnames(counts3))
design_matrix_bred <- model.matrix( ~ Condition2[Bred_cols])
colnames(design_matrix_bred)
design_matrix_bred
y_for_DE_bred <- DGEList(counts=counts3[, Bred_cols])
y_for_DE_bred <- calcNormFactors(y_for_DE_bred)
y_for_DE_bred = estimateDisp(y_for_DE_bred, design=design_matrix_bred) 
plotBCV(y_for_DE_bred)
fit_bred <- glmFit(y_for_DE_bred, design=design_matrix_bred, dispersion=y_for_DE_bred$trended.dispersion)
```

Virus: KO vs WT
```{r}
lrt_virus <- glmLRT(fit_virus,coef = "Condition2[Virus_cols]KO")
topTags(lrt_virus)
smoothScatter(x=lrt_virus$table$logFC, y=-log10(lrt_virus$table$PValue))
lrt_virus_table = data.frame(lrt_virus$table, p.adjust(lrt_virus$table$PValue, method = "BH"))
colnames(lrt_virus_table) = c(colnames(lrt_virus$table), "adjPVal_BH")
hist(lrt_virus_table$PValue)
```

Bred: KO vs WT
```{r}
lrt_bred <- glmLRT(fit_bred,coef = "Condition2[Bred_cols]KO")
topTags(lrt_bred)
smoothScatter(x=lrt_bred$table$logFC, y=-log10(lrt_bred$table$PValue))
lrt_bred_table = data.frame(lrt_bred$table, p.adjust(lrt_bred$table$PValue, method = "BH"))
colnames(lrt_bred_table) = c(colnames(lrt_bred$table), "adjPVal_BH")
hist(lrt_bred_table$PValue)
```
Format output into excel sheet. 
```{r}
genes = rownames(counts3)
ensembl = useEnsembl(biomart="ensembl", dataset="mmusculus_gene_ensembl")
head(listAttributes(ensembl), 30)
mmusculus_swissprot <- getBM(attributes=c('ensembl_gene_id','external_gene_name','description'),filters = 'ensembl_gene_id',  values = genes , mart = ensembl)

#fpkm_log_matrix2 = rpkm (y_for_DE_virus, gene.length=counts1[keep,"GeneLength"], log = TRUE)
index = match (genes, mmusculus_swissprot$ensembl_gene_id)
results = data.frame(mmusculus_swissprot[index,], counts3, fpkm_log_matrix, lrt_virus_table, lrt_bred_table)
colnames(results) = c(
                    colnames(mmusculus_swissprot[index,])
                  , unlist(lapply (colnames(counts3), function(x) paste(x, "counts", sep="_")))
                  , unlist(lapply (colnames(fpkm_log_matrix), function(x) paste(x, "fpkm", sep="_")))
                  , unlist(lapply (colnames(lrt_virus_table), function(x) paste(x, "_virus_KO_vs_WT_baseline", sep="_")))
                  , unlist(lapply (colnames(lrt_bred_table), function(x) paste(x, "_bred_KO_vs_WT_baseline", sep="_"))) )
curr_date = format(Sys.time(), "%a_%b_%d_%H_hrs_%M_min_%S_sec_%Y")
write.table(results, paste0("~/Documents/chang/chang_", curr_date, ".xls"), row.names=F, col.names=T, sep="\t")

```

Look at agreement between different analyses. 

My analysis (edgeR): 
KO_vs_WT_virus1 
KO_vs_WT_bred1

NovoGene's analysis: 
KO_vs_WT_virus2 
KO_vs_WT_bred2 
```{r}

library(UpSetR)
upsetr_matrix  = data.frame(cbind(as.integer(results$adjPVal_BH__virus_KO_vs_WT_baseline<0.05), as.integer(results$adjPVal_BH__bred_KO_vs_WT_baseline<0.05), as.integer(counts1$padj.KO_virusvsWT_virus.< 0.05)[keep], as.integer(counts1$padj.KO_bredvsWT_bred. < 0.05)[keep]))
colnames(upsetr_matrix) = c ("KO_vs_WT_virus1", "KO_vs_WT_bred1", "KO_vs_WT_virus2", "KO_vs_WT_bred2")
#remove rows with one or more NA 
upsetr_matrix2 = upsetr_matrix[which(complete.cases(upsetr_matrix)), ]
dim(upsetr_matrix)
dim(upsetr_matrix2)
colSums(upsetr_matrix2)
upset(upsetr_matrix2, sets = c("KO_vs_WT_virus1", "KO_vs_WT_bred1",  "KO_vs_WT_virus2", "KO_vs_WT_bred2"),sets.bar.color = "#56B4E9", order.by = "freq", empty.intersections = "on")

```
Quick sanity check.  
```{r}
#compare fpkm values  check
qqplot(results$Bred_WT_rep1_counts, counts1$WT73M_readcount[keep])
cor(results$Bred_WT_rep1_counts, counts1$WT73M_readcount[keep])
cor(results$Bred_WT_rep1_fpkm, counts1$WT73M_fpkm[keep])
cor(results$Virus_WT_rep2_fpkm, counts1$WT113M_fpkm[keep])
cor(results$Virus_KO_rep2_fpkm, counts1$KO117M_fpkm[keep])

```
Compare using Venn diagram. 
```{r}
library(limma)
venn1 = vennCounts(upsetr_matrix2[,c("KO_vs_WT_virus1", "KO_vs_WT_virus2" )])
vennDiagram(venn1)

venn2 = vennCounts(upsetr_matrix2[,c("KO_vs_WT_bred1", "KO_vs_WT_bred2")])
vennDiagram(venn2)
```
Try to use DEseq2 package(because that is what NovoGene used) : Virus KO vs WT 
```{r}
library("DESeq2")
dds_virus <- DESeqDataSetFromMatrix(countData = counts3[, Virus_cols],
                              colData = cold [Virus_cols,],
                              design = ~ Condition2)
dds_virus
dds_virus <- DESeq(dds_virus)
res_virus <- results(dds_virus)
res_virus
hist(res_virus$pvalue)
```
Try to use DEseq2 : Bred KO vs WT
```{r}
dds_bred <- DESeqDataSetFromMatrix(countData = counts3[, Bred_cols],
                              colData = cold [Bred_cols,],
                              design = ~ Condition2)
dds_bred
dds_bred <- DESeq(dds_bred)
res_bred <- results(dds_bred)
res_bred
hist(res_bred$pvalue)
```
The two DESeq2 analyses agree more which makes perfect sense. 
My analysis (edgeR): 
KO_vs_WT_virus1 
KO_vs_WT_bred1

NovoGene's analysis: 
KO_vs_WT_virus2 
KO_vs_WT_bred2 

My analysis (DESeq2):
KO_vs_WT_virus3
KO_vs_WT_bred3

```{r}

library(UpSetR)
upsetr_matrix  = data.frame(cbind(as.integer(results$adjPVal_BH__virus_KO_vs_WT_baseline<0.05), as.integer(results$adjPVal_BH__bred_KO_vs_WT_baseline<0.05), as.integer(counts1$padj.KO_virusvsWT_virus.< 0.05)[keep], as.integer(counts1$padj.KO_bredvsWT_bred. < 0.05)[keep], as.integer(res_virus$padj<0.05), as.integer(res_bred$padj<0.05)))
colnames(upsetr_matrix) = c ("KO_vs_WT_virus1", "KO_vs_WT_bred1", "KO_vs_WT_virus2", "KO_vs_WT_bred2", "KO_vs_WT_virus3", "KO_vs_WT_bred3" )
#remove rows with one or more NA 
upsetr_matrix2 = upsetr_matrix[which(complete.cases(upsetr_matrix)), ]
dim(upsetr_matrix)
dim(upsetr_matrix2)
colSums(upsetr_matrix2)
upset(upsetr_matrix2, sets = c("KO_vs_WT_virus1", "KO_vs_WT_bred1",  "KO_vs_WT_virus2", "KO_vs_WT_bred2","KO_vs_WT_virus3", "KO_vs_WT_bred3"),sets.bar.color = "#56B4E9", order.by = "freq", empty.intersections = "on")

venn1 = vennCounts(upsetr_matrix2[,c("KO_vs_WT_virus1", "KO_vs_WT_virus2", "KO_vs_WT_virus3" )])
vennDiagram(venn1)

venn2 = vennCounts(upsetr_matrix2[,c("KO_vs_WT_bred1", "KO_vs_WT_bred2", "KO_vs_WT_bred3")])
vennDiagram(venn2)



venn3 = vennCounts(upsetr_matrix2[,c( "KO_vs_WT_virus2", "KO_vs_WT_virus3" )])
vennDiagram(venn3)

venn4 = vennCounts(upsetr_matrix2[,c( "KO_vs_WT_bred2", "KO_vs_WT_bred3")])
vennDiagram(venn4)

```