---
title: "Endoderm dox vs no treatment hg38"
output: html_document
---

```{r setup, include=FALSE}
library(pheatmap)
library(edgeR)
library(DESeq2)
library(biomaRt)
library(ggplot2)
library(topGO)
require(plyr)
library(GOplot)
library("org.Hs.eg.db")
sessionInfo()
```
Read in file with read counts.
```{r cars}
filename = "/Users/tfriedrich//Box Sync/laptop_folders//Mattis/endoderm_project/readcounts.txt"
counts1 = read.table(filename, header = T, stringsAsFactors = F) 
rownames(counts1) = counts1$Geneid

```
Reformat.
```{r pressure, echo=FALSE}
counts2 = counts1[,7:ncol(counts1)]
colnames(counts2) = paste( matrix(unlist(strsplit(colnames(counts2), "\\.")), byrow =T, ncol=7)[,3], sub("_1_algn_Aligned", "", matrix(unlist(strsplit(colnames(counts2), "\\.")), byrow =T, ncol=7)[,4]), sep="-")
colnames(counts2)
```
More reformatting.
```{r}
rename = read.table("/Users/tfriedrich/Box Sync/laptop_folders//Mattis/endoderm_project/metadata_Aras_Mattis_2018_08_01v3_Endoderm_dox_config.txt", fill=T, header=T, sep="\t")

renamesample=list()
for (x in 1:nrow(rename)){
  oldname = sub("_1.fq.gz", "", rename$Location[x])
  print (rename$cell.type[x])  #check to make sure every sample is endoderm 
  newname = paste(rename$Name[x], rename$dox.treatment[x], rename$Genotype[x], sep="_")
  renamesample[oldname] = newname
}
colnames(counts2) = unlist(lapply (colnames(counts2), function(x) renamesample[[x]]))

```
Filter out genes that are too highly expressed or lowly expressed. 
```{r}
keep = rowMeans(counts2) > 5 & rowMeans(counts2) < 5000 
counts3 = counts2[keep,]
```
Take the RPKM.
```{r}
y <- DGEList(counts=counts3)
fpkm_matrix = rpkm (y, gene.length=counts1[keep,"Length"], log = FALSE)
```
Dendrogram showing relationships using fpkm. 
```{r}
#par(cex=1,font=3)
hc2 <- hclust(stats::dist(t(fpkm_matrix), method="minkowski"), "ward.D2")
#par(mar=c(10, 4, 4, 10), pty='s')
plot(hc2, lwd=4, lty=1, col="black", col.lab="red", xlab="Samples", main="Samples clustered by FPKM using Ward's clustering & Euclidean distance")
```

PCA
```{r}
Condition1 = factor(matrix (unlist(strsplit(colnames(counts3), "_")), byrow=T, ncol=3)[,2], levels = c("No", "Yes"))
Condition2 = factor(matrix (unlist(strsplit(colnames(counts3), "_")), byrow=T, ncol=3)[,3], levels = c("wild-type","NASH pt"))
Condition3 = factor(matrix (unlist(strsplit(colnames(counts3), "_")), byrow=T, ncol=3)[,1])

cold <- data.frame("Condition1"=Condition1, "Condition2"=Condition2, "Condition3"=Condition3) 
rownames(cold) = colnames(counts3)
cold

ddsMat <- DESeq2::DESeqDataSetFromMatrix(countData=counts3, colData=cold, design=~Condition1 + Condition3 );
colnames(ddsMat) <- colnames(counts3)
rld <- DESeq2::rlog(ddsMat) # LOG TRANSFORMED.

```

```{r}
par(mfrow=c(2,1))
pca <- prcomp(t(assay(rld)))

d <- data.frame(pca$x, Condition1, Condition2,  name=colnames(rld)) # returns PC1, PC2, ... PC8... etc
ppp <- ggplot(d, aes_string(x="PC1", y="PC2",shape="Condition1", color="Condition2"))
ppp <- ppp + geom_point(size=6)
plot(ppp)

d <- data.frame(pca$x, group=Condition1, Condition3,  name=colnames(rld)) # returns PC1, PC2, ... PC8... etc
ppp <- ggplot(d, aes_string(x="PC1", y="PC2",shape="Condition1", color="Condition3"))
ppp <- ppp + geom_point(size=6)
plot(ppp)
```
Some of the related samples are messing up clustering. 

Heatmap to see how cells cluster by cell type or experimental condition. 
```{r}
HEATMAP_HEIGHT <- 45 # inchescol
HEATMAP_WIDTH  <- 15
annotation <- data.frame(
  Condition1= Condition1, Condition2=Condition2)
rownames(annotation) <- colnames(fpkm_matrix)
pheatmap(fpkm_matrix,
         , width=HEATMAP_WIDTH, height=HEATMAP_HEIGHT
         , scale="row"  # actually scale each row
         , clustering_distance_rows="correlation" # or: euclidean
         , clustering_distance_cols="correlation"
         , clustering_method="complete"
         , display_numbers=FALSE
         , border_color="#00000000" # note: the last two digits here are the alpha channel/ transparency
         , show_rownames=FALSE, show_colnames = TRUE
         , annotation = annotation
         , fontsize_row=3.5)
```

Fit the model and estimate dispersion value.
```{r}

design_matrix <- model.matrix( ~ Condition1 + Condition2 )
colnames(design_matrix)
design_matrix
y_for_DE <- DGEList(counts=counts3)
y_for_DE <- calcNormFactors(y_for_DE)
y_for_DE = estimateDisp(y_for_DE, design=design_matrix) 
plotBCV(y_for_DE)
fit <- glmFit(y_for_DE, design=design_matrix, dispersion=y_for_DE$trended.dispersion)
```
Pretty high estimated dispersion values. 


Tx vs no tx (dox)
```{r}
lrt <- glmLRT(fit, coef = "Condition1Yes")
topTags(lrt)
smoothScatter(x=lrt$table$logFC, y=-log10(lrt$table$PValue))
```
More genes appear to be down-regulated. 

```{r}
lrt_table = data.frame(lrt$table, p.adjust(lrt$table$PValue, method = "BH"))
colnames(lrt_table) = c(colnames(lrt$table), "adjPVal_BH")
hist(lrt_table$PValue)
```


Format output into excel sheet. 
```{r}
genes = rownames(counts3)
ensembl = useEnsembl(biomart="ensembl", dataset="hsapiens_gene_ensembl")
head(listAttributes(ensembl), 30)
hgnc_swissprot <- getBM(attributes=c('ensembl_gene_id','hgnc_symbol','description'),filters = 'ensembl_gene_id',  values = genes , mart = ensembl)

index = match (genes, hgnc_swissprot$ensembl_gene_id)
results = data.frame(hgnc_swissprot[index,], counts3, fpkm_matrix, lrt_table)
colnames(results) = c(
                    colnames(hgnc_swissprot[index,])
                  , unlist(lapply (colnames(counts3), function(x) paste(x, "counts", sep="_")))
                  , unlist(lapply (colnames(fpkm_matrix), function(x) paste(x, "fpkm", sep="_")))
                  , unlist(lapply (colnames(lrt_table), function(x) paste(x, "_dox_vs_nodox_baseline", sep="_")))
                  )
curr_date = format(Sys.time(), "%a_%b_%d_%H_hrs_%M_min_%S_sec_%Y")
write.table(results, paste0("~/Box Sync/laptop_folders//Mattis/endoderm_project/mattis_", curr_date, ".xls"), row.names=rownames(counts3), col.names=NA, sep="\t")

```
cell death genes
```{r}
go1 = read.table("~/Downloads/GO-0008219.txt", sep="\t")
dim(go1)

go1[, "gene"] = sub ("UniProtKB:", "", go1$V1)

#results_go1 = results[which(results$hgnc_symbol %in% go1$gene),]

#convert uniprot to gene symbol
annotation <- getBM(attributes=c("uniprotswissprot", "hgnc_symbol", "uniprot_gn"), filters="uniprotswissprot", values=go1[, "gene"], mart=ensembl)

index_go1 = match ( go1$gene, annotation$uniprot_gn)
go1_gn = cbind( annotation[index_go1, ], go1 )

results_go1 = results[which(results$hgnc_symbol %in% go1_gn$hgnc_symbol[!is.na(go1_gn$hgnc_symbol)]),]
dim(results_go1)

HEATMAP_HEIGHT <- 45 # inchescol
HEATMAP_WIDTH  <- 15
#annotation <- data.frame(
#  Condition1= Condition1, Condition2=Condition2)
#rownames(annotation) <- colnames(fpkm_matrix)
pdf("~/Box Sync/laptop_folders/Mattis/endoderm_project/GO-0008219.pdf")
pheatmap(results_go1[,grep ("_fpkm", colnames(results_go1))],
         , width=HEATMAP_WIDTH, height=HEATMAP_HEIGHT
         , scale="row"  # actually scale each row
         #, clustering_distance_rows="correlation" # or: euclidean
         , clustering_distance_cols="correlation"
         , clustering_method="complete"
         , display_numbers=FALSE
        , border_color="#00000000" # note: the last two digits here are the alpha channel/ transparency
         , show_rownames=FALSE, show_colnames = TRUE
         #, annotation = annotation
         , fontsize_row=3.5)
dev.off()

```


```{r}
#setup gene DE for GOplot package later
GOplot_genes = results [grep ("__dox_vs_nodox_baseline",colnames(results))]
colnames(GOplot_genes) = c("logFC", "logCPM", "LR", "Pvalue", "BH")
GOplot_genes["ID"] = results[,"ensembl_gene_id"]
colnames(GOplot_genes)
length(unique(GOplot_genes$ID))
GOplot_genes = GOplot_genes[, c("ID", "logFC")]

geneIDs = results[,"ensembl_gene_id"]
inUniverse = geneIDs %in%  results[,"ensembl_gene_id"]
inSelection = geneIDs %in% results[which(results[,"adjPVal_BH__dox_vs_nodox_baseline"] <=0.05),"ensembl_gene_id"]
alg <- factor( as.integer( inSelection[inUniverse] ) )
names(alg) <- geneIDs[inUniverse]
head(alg)

onts = c( "MF", "BP", "CC" )
tab = as.list(onts)
names(tab) = onts
GOplot_terms=list()
  for(i in 1:3) {
    ## prepare data
    tgd <- new( "topGOdata", ontology=onts[i], allGenes = alg, nodeSize=5,
                 annot=annFUN.org, mapping="org.Hs.eg.db", ID = "ensembl" )

    ## run tests
    resultTopGO.elim <- runTest(tgd, algorithm = "elim", statistic = "Fisher" )
    resultTopGO.classic <- runTest(tgd, algorithm = "classic", statistic = "Fisher" )

    ## look at results
    tab[[i]] <- GenTable( tgd, Fisher.elim = resultTopGO.elim, 
        Fisher.classic = resultTopGO.classic,
        orderBy = "Fisher.classic" , topNodes = 200)

    #convert list to dataframe  
    allGO = genesInTerm(tgd, tab[[i]][,"GO.ID"])
    #allGO_genes = ldply(allGO,data.frame)[,2]
    #length(unique(allGO_genes))
    #GOplot_genes = GOplot_genes[GOplot_genes$ID %in% allGO_genes,]
    test = ldply (lapply(allGO, function(x) paste(x, collapse=",")), data.frame)
    #match(test$.id, tab[[i]][, "GO.ID"]) #check to make sure they rows match 
    GOplot_terms[[i]] = data.frame(onts[i], tab[[i]], test, stringsAsFactors = FALSE)
    GOplot_terms[[i]][, "Fisher.elim"] = as.numeric(as.character(GOplot_terms[[i]][, "Fisher.elim"]))
    GOplot_terms[[i]][, "Fisher.elim.BH"] = p.adjust(GOplot_terms[[i]][, "Fisher.elim"], method="BH")
    GOplot_terms[[i]] = GOplot_terms[[i]][, c("onts.i.", "GO.ID", "Term", "Fisher.elim.BH", "X..i..")]
    colnames(GOplot_terms[[i]]) = c("Category", "ID", "Term",  "adj_pval", "Genes")
  }

topGOResults <- rbind.fill(tab)
write.csv(topGOResults, file = "topGOResults.csv")
topGOResults2 <- rbind.fill(GOplot_terms)
write.csv(topGOResults2, file = "topGOResults2.csv")


circ <- circle_dat(topGOResults2, data.frame(GOplot_genes))
GOBar(circ)
GOBar(circ, display='multiple')
GOBubble(circ, labels = 1.5)
GOCircle(circ)

genesInNegProlif =genesInTerm(new( "topGOdata", ontology=onts[3], allGenes = alg, nodeSize=5,
                 annot=annFUN.org, mapping="org.Mm.eg.db", ID = "ensembl" ), "GO:0008285")
```


```{r}

library(plotly)
packageVersion('plotly')
m <- as.matrix(results_go1[1:100,grep ("_fpkm", colnames(results_go1))])
p <- plot_ly(
    #x = c("a", "b", "c"), y = c("d", "e", "f"),
    y = rownames(results_go1[1:100,grep ("_fpkm", colnames(results_go1))]), x = colnames(results_go1[1:100,grep ("_fpkm", colnames(results_go1))]), 
    z = m, type = "heatmap"
)
# Create a shareable link to your chart
# Set up API credentials: https://plot.ly/r/getting-started
chart_link = api_create(p, filename="heatmap-cat")
chart_link


Sys.setenv("plotly_username"="tara.friedrich")
Sys.setenv("plotly_api_key"="0VBILd5Cg74iqNBG8aEl")
```
