---
title: "2018_09_12_Chang_Weng"
output: html_document
---
```{r}
library(pheatmap)
library(edgeR)
library(DESeq2)
library(biomaRt)
library(ggplot2)
sessionInfo()
```

Friday, September 7, 2018 
WT73M_readcount               Bred WT
WT58M_readcount               Bred WT
WT973M_readcount             Bred WT
KO87M_readcount                Bred KO
KO56M_readcount                Bred KO
KO972M_readcount              Bred KO
WT112M_readcount             Virus WT
WT113M_readcount             Virus WT
WT114M_readcount             Virus WT
KO116M_readcount              Virus KO
KO117M_readcount              Virus KO
KO119M_readcount              Virus KO



```{r }
#library(readxl)
setwd("~/Downloads/chang/")
#originaldata = read_excel("gene_list/gene.description.xls", sheet="gene.description") #disappointed this didn't work
                         #need to download the developer version to see if that works
#devtools::install_version("readxl", version = "1.0.0")
#https://github.com/tidyverse/readxl/issues/466

```
```{r}
filename = "readcounts.txt"
counts1 = read.table(filename, header = T, stringsAsFactors = F) 
rownames(counts1) = counts1$Geneid

```
Remove metadata (Gene, coordinates, etc)
```{r}
counts2 = counts1[,grep("readcount", colnames(counts1))]
sample_names = list(
  "WT73M_readcount" = "Bred_WT_rep1"
, "WT58M_readcount" = "Bred_WT_rep2"
, "WT973M_readcount" = "Bred_WT_rep3"
, "KO87M_readcount" = "Bred_KO_rep1"
, "KO56M_readcount" = "Bred_KO_rep2"
, "KO972M_readcount" = "Bred_KO_rep3"
, "WT112M_readcount" = "Virus_WT_rep1"
, "WT113M_readcount" = "Virus_WT_rep2"
, "WT114M_readcount" = "Virus_WT_rep3"
, "KO116M_readcount" = "Virus_KO_rep1"
, "KO117M_readcount" = "Virus_KO_rep2"
, "KO119M_readcount" = "Virus_KO_rep3"
)
colnames(counts2) = unlist(lapply (colnames(counts2), function(x) sample_names[[x]]))
colnames(counts2)
```
Filter out genes with low and high counts ( high counts might be mitochondrial genes)
```{r}
keep = rowMeans(counts2) > 5 & rowMeans(counts2) < 5000 
counts3 = counts2[keep,]
dim(counts2)
dim(counts3)
```

Take the log (RPKM)
```{r}
y <- DGEList(counts=counts3)
fpkm_log_matrix = rpkm (y, gene.length=counts1[keep,"GeneLength"], log = TRUE)
```
Dendrogram 
```{r}
# 
#par(cex=1,font=3)
hc2 <- hclust(stats::dist(t(fpkm_log_matrix), method="minkowski"), "ward.D2")
#par(mar=c(10, 4, 4, 10), pty='s')
plot(hc2, lwd=4, lty=1, col="black", col.lab="red", xlab="Samples", main="Samples clustered by log2(FPKM+1) using Ward's clustering & Euclidean distance")
```
```{r}
Condition1 = factor(matrix (unlist(strsplit(colnames(counts3), "_")), byrow=T, ncol=3)[,1], levels = c("Bred", "Virus"))
Condition2 = factor(matrix (unlist(strsplit(colnames(counts3), "_")), byrow=T, ncol=3)[,2], levels = c("KO", "WT"))
Condition2 = relevel(Condition2, ref="WT")
cold <- data.frame("Condition1"=Condition1, "Condition2"=Condition2) 
cold

ddsMat <- DESeq2::DESeqDataSetFromMatrix(countData=counts3, colData=cold, design=~Condition1 + Condition2);
colnames(ddsMat) <- colnames(counts3)
rld <- DESeq2::rlog(ddsMat) # LOG TRANSFORMED.
#rld <- DESeq2::rlog(ddsMat, blind=blind_status) 
```

```{r}
pca              <- prcomp(t(assay(rld)))
#barplot(allPer.vec, main="PCA: % variance explained by each component\n" , ylim=c(0,100), names.arg=allPerLabels.vec)

d <- data.frame(pca$x, group=Condition1, Condition2, name=colnames(rld)) # returns PC1, PC2, ... PC8... etc
ppp <- ggplot(d, aes_string(x="PC1", y="PC2", shape="Condition2", color="group"))
ppp <- ppp + geom_point(size=6)
plot(ppp)
```
Heatmap to see how cells cluster by cell type or experimental condition.  No replicates because they have been combined. 
```{r}
HEATMAP_HEIGHT <- 45 # inchescol
HEATMAP_WIDTH  <- 15
annotation <- data.frame(
  Condition1= Condition1, Condition2=Condition2)
rownames(annotation) <- colnames(fpkm_log_matrix)
pheatmap(fpkm_log_matrix,
         , width=HEATMAP_WIDTH, height=HEATMAP_HEIGHT
         , scale="row"  # actually scale each row
         , clustering_distance_rows="correlation" # or: euclidean
         , clustering_distance_cols="correlation"
         , clustering_method="complete"
         , display_numbers=FALSE
         , border_color="#00000000" # note: the last two digits here are the alpha channel/ transparency
         , show_rownames=FALSE, show_colnames = TRUE
         , annotation = annotation
         , fontsize_row=3.5)
```

Fit the model and estimate dispersion value.
```{r}
Virus_cols = grep ("Virus", colnames(counts3))
design_matrix_virus <- model.matrix( ~ Condition2[Virus_cols])
colnames(design_matrix_virus)
design_matrix_virus
y_for_DE_virus <- DGEList(counts=counts3[, Virus_cols])
y_for_DE_virus <- calcNormFactors(y_for_DE_virus)
y_for_DE_virus = estimateDisp(y_for_DE_virus, design=design_matrix_virus) 
plotBCV(y_for_DE_virus)
fit_virus <- glmFit(y_for_DE_virus, design=design_matrix_virus, dispersion=y_for_DE_virus$trended.dispersion)
```
```{r}
Bred_cols = grep ("Bred", colnames(counts3))
design_matrix_bred <- model.matrix( ~ Condition2[Bred_cols])
colnames(design_matrix_bred)
design_matrix_bred
y_for_DE_bred <- DGEList(counts=counts3[, Bred_cols])
y_for_DE_bred <- calcNormFactors(y_for_DE_bred)
y_for_DE_bred = estimateDisp(y_for_DE_bred, design=design_matrix_bred) 
plotBCV(y_for_DE_bred)
fit_bred <- glmFit(y_for_DE_bred, design=design_matrix_bred, dispersion=y_for_DE_bred$trended.dispersion)
```

Virus: KO vs WT
```{r}
lrt_virus <- glmLRT(fit_virus,coef = "Condition2[Virus_cols]KO")
topTags(lrt_virus)
smoothScatter(x=lrt_virus$table$logFC, y=-log10(lrt_virus$table$PValue))
lrt_virus_table = data.frame(lrt_virus$table, p.adjust(lrt_virus$table$PValue, method = "BH"))
colnames(lrt_virus_table) = c(colnames(lrt_virus$table), "adjPVal_BH")
hist(lrt_virus_table$PValue)
```

Bred: KO vs WT
```{r}
lrt_bred <- glmLRT(fit_bred,coef = "Condition2[Bred_cols]KO")
topTags(lrt_bred)
smoothScatter(x=lrt_bred$table$logFC, y=-log10(lrt_bred$table$PValue))
lrt_bred_table = data.frame(lrt_bred$table, p.adjust(lrt_bred$table$PValue, method = "BH"))
colnames(lrt_bred_table) = c(colnames(lrt_bred$table), "adjPVal_BH")
hist(lrt_bred_table$PValue)
```
Format output into excel sheet. 
```{r}
genes = rownames(counts3)
ensembl = useEnsembl(biomart="ensembl", dataset="mmusculus_gene_ensembl")
head(listAttributes(ensembl), 30)
mmusculus_swissprot <- getBM(attributes=c('ensembl_gene_id','hgnc_symbol','description'),filters = 'ensembl_gene_id',  values = genes , mart = ensembl)

#fpkm_log_matrix2 = rpkm (y_for_DE_virus, gene.length=counts1[keep,"GeneLength"], log = TRUE)
index = match (genes, mmusculus_swissprot$ensembl_gene_id)
results = data.frame(mmusculus_swissprot[index,], counts3, fpkm_log_matrix, lrt_virus_table, lrt_bred_table)
colnames(results) = c(
                    colnames(mmusculus_swissprot[index,])
                  , unlist(lapply (colnames(counts3), function(x) paste(x, "counts", sep="_")))
                  , unlist(lapply (colnames(fpkm_log_matrix), function(x) paste(x, "fpkm", sep="_")))
                  , unlist(lapply (colnames(lrt_virus_table), function(x) paste(x, "_virus_KO_vs_WT_baseline", sep="_")))
                  , unlist(lapply (colnames(lrt_bred_table), function(x) paste(x, "_bred_KO_vs_WT_baseline", sep="_"))) )
curr_date = format(Sys.time(), "%a_%b_%d_%H_hrs_%M_min_%S_sec_%Y")
write.table(results, paste0("~/Downloads/chang/chang_", curr_date, ".xls"), row.names=F, col.names=T, sep="\t")

```
